---
title: "Lesson 2: Quarto and Reproducible Reporting"
---

## Learning goals

By the end of this lesson, you will be able to:

- Explain what Quarto is and why it is used in public health research
- Create and render a Quarto document
- Use Markdown to structure writing
- Insert and run R code chunks
- Use inline R code to report values reproducibly

---

# What is Quarto?

Quarto is a tool for creating **reproducible documents** that combine writing and data analysis in one place. A Quarto document can include text, code, results, tables, and figures so that your report updates automatically when your analysis changes.

In public health, Quarto supports transparency and accuracy by keeping your analysis and your writing connected.

In this course, all homework assignments and projects will be written in Quarto.

---

## What does "reproducible" mean?

A reproducible document is one where another person (or future you) can:

- open the file
- run the code
- get the same results
- see exactly how the results were produced

Reproducibility helps prevent errors that happen when people copy and paste numbers or figures from the Console into a report.

---

## Anatomy of a Quarto document

A Quarto document typically has three pieces:

1. A YAML header (at the very top)
2. Written text (formatted using Markdown)
3. Code chunks (R code that runs when you render)

You will see all three in this lesson.

---

## Creating a Quarto document in RStudio

To create a new Quarto document:

1. Go to File → New File → Quarto Document
2. Choose HTML as the output format
3. Give the file a descriptive name
4. Click Create

To render:

- Click the Render button, or
- Use Quarto Render from the Build tab (depending on your setup)

:::{.callout-tip}
When you render, Quarto runs the code chunks from top to bottom and builds the final output.
:::
---

## Markdown: formatting your writing

Markdown is a simple way to format text. In this course, Markdown will be used for headings, lists, emphasis, links, and section organization.

### Headings

Headings help organize reports. In Markdown, headings use the number sign.

Example:

    # Main section
    ## Subsection
    ### Smaller subsection

### Bold and italics

Example:

    **bold text**
    *italic text*

### Lists

Example:

    - first item
    - second item
    - third item

Markdown keeps your writing clean and readable.

---

## Code chunks: where analysis happens

R code is written inside code chunks. Code chunks run when you render the document.

Here is a simple code chunk:

```{r}
2 + 2
```

---

## Running chunks vs rendering the document

You can run code in two ways:

- Run a single chunk using the green Run button inside the chunk
- Render the entire document to produce the final report

In this course, you should always **render** your document before submitting so your outputs are up to date.

---

## A simple data example

Below we create a small vector and compute summary statistics.

```{r}
ages <- c(21, 34, 29, 45, 38, 41)

mean(ages)
sd(ages)
length(ages)
```

---

## Inline R code

Inline code allows you to insert R results directly into a sentence. This is one of the most important skills in the course because it prevents copy and paste errors.

Inline code looks like this inside a sentence:

The sample included `r length(ages)` participants, and the mean age was `r round(mean(ages), 1)` years.

When you render, Quarto replaces the inline code with the calculated value.

---

## Why inline code matters in public health writing

Inline code helps you:

- keep your writing consistent with your analysis
- update values automatically when your data change
- reduce mistakes when reporting results

In this course, students are expected to use inline code for reported statistics whenever possible, especially in Projects 2 through 4.

---

## Showing or hiding code

Sometimes you want to show the results but hide the code (for a polished report). You can hide code using an option at the top of the chunk.

```{r}
#| echo: false
mean(ages)
```

In the rendered document, the code will be hidden and only the result will appear.

---

## Common mistakes to avoid

Avoid these common issues:

- Typing numeric results manually into your report instead of using inline code
- Copying results from the Console into your writing
- Forgetting to render after changing code or data
- Using inline code that refers to an object that has not been created yet

A simple rule is: if a number appears in your report, ask yourself whether it was produced by R during rendering.

---

## Course expectations for Quarto work

In this course, students are expected to:

- complete homework and projects in Quarto
- organize work using headings and clear sections
- write interpretations in complete sentences
- use inline code for key reported numbers when possible
- render before submitting

Perfection is not expected. Consistent effort and honest engagement are the priority.

---

## Weekly assignment reminder

Students will:

- Create their first Quarto document
- Add headings, text, and code chunks
- Include at least one chunk that prints a small dataset or summary
- Write 1–2 paragraphs explaining what reproducibility means in science

Goal: Establish documentation habits early

---

## Practice section

Create a new section in your own homework file and complete the tasks below.

1. Create a numeric vector with at least 6 values.
2. Use inline code to report:
   - sample size
   - mean
   - standard deviation
3. Change one value and re-render to confirm the sentence updates.

Here is a starter example you may adapt:

```{r}
x <- c(4, 6, 9, 12, 15, 18)
```

Write a sentence like:

The sample size was `r length(x)`, the mean was `r round(mean(x), 2)`, and the standard deviation was `r round(sd(x), 2)`.

---

## Key takeaways

Quarto allows you to combine writing and analysis. Markdown organizes your text. Code chunks run R. Inline code inserts calculated results into sentences. Reproducibility is a core skill for public health data science.

---

## Looking ahead

Next week, we will begin data visualization using ggplot2 and you will continue writing all work in Quarto.
