---
title: "Lesson 2 Review: R Workflow, Quarto, and Core Functions"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose of this review lesson

This lesson reviews and reinforces the foundational material from **Lessons 1 and 2**. Before moving on to visualization and data wrangling, it is important that you understand:

- how R scripts differ from Quarto documents  
- how RStudio, Quarto, code chunks, and inline code fit together  
- the core workflow expected in this course  
- essential R functions you will use repeatedly  

This lesson is designed as a reference you can return to throughout the semester.

---

## R workflow overview

In this course, your workflow follows this pattern:

1. Write code in R scripts or Quarto documents  
2. Use R to create objects, summaries, figures, and models  
3. Use Quarto to combine analysis and writing  
4. Render documents to produce final reports  

Understanding when to use **a script** versus **a Quarto document** is critical.

---

## R scripts vs Quarto documents

### R scripts (`.R` files)

An R script is a plain text file that contains **only R code**.

You use R scripts to:

- write and test code  
- clean and prepare data  
- experiment with analyses  
- run code step by step  

R scripts do **not** include formatted writing, figures with captions, or narrative interpretation.

Example use case:
You are cleaning a dataset or testing code before putting results into a report.

---

### Quarto documents (`.qmd` files)

A Quarto document combines **writing and R code** in a single file.

You use Quarto documents to:

- write homework assignments  
- create reports and manuscripts  
- integrate code, figures, tables, and interpretation  
- produce reproducible outputs  

A Quarto document includes:
- a YAML header  
- written text (Markdown)  
- R code chunks  
- inline R code  

In this course, **all submitted work must be written in Quarto** unless otherwise stated.

---

## Key difference (important)

- Use **R scripts** to *develop and test* code  
- Use **Quarto documents** to *present and submit* work  

Most students will use **both**:
Write code in scripts → move finalized code into Quarto.

---

## Review: Week 1 concepts (R basics)

From Week 1, you should understand:

### RStudio interface

- Source: where scripts and Quarto documents are written  
- Console: where R runs code  
- Environment: where objects are stored  
- Files/Plots/Help: where outputs appear  

You should primarily work in the **Source pane**, not the Console.

---

### Objects and assignment

You create objects using `<-`.

```{r}
x <- 10
y <- 5
```

Objects store values so you can reuse them.

---

### Vectors

Vectors store multiple values of the same type.

```{r}
ages <- c(23, 35, 41, 29, 50)
```

Common functions for vectors:

```{r}
mean(ages)
median(ages)
sd(ages)
length(ages)
```

---

### Functions

Functions take inputs and return outputs.

```{r}
mean(ages)
```

Understanding how to use functions is more important than memorizing them.

---

### Packages

Packages extend R.

You install once:

```{r}
# install.packages("tidyverse")
```

You load every session:

```{r}
library(tidyverse)
```

---

## Review: Week 2 concepts (Quarto & reproducibility)

### YAML header

Every Quarto document begins with a YAML header.

```yaml
---
title: "My Document"
format: html
---
```

The YAML controls document settings.

---

### Markdown

Markdown formats your writing using simple syntax.

You use it for:
- headings  
- lists  
- emphasis  
- document structure  

Markdown keeps reports readable and professional.

---

### Code chunks

R code is written inside code chunks.

```{r}
2 + 2
```

Code chunks run when you **render** the document.

---

### Rendering

Rendering:
- runs all code chunks  
- generates the final document  
- updates all figures and inline values  

Always render before submitting.

---

### Inline R code

Inline code inserts R results into sentences.

```{r}
ages <- c(23, 35, 41, 29, 50)
```

The sample included `r length(ages)` participants, and the mean age was `r round(mean(ages), 1)` years.

Inline code prevents copy-and-paste errors and is required for reporting values in this course.

---

## Go-to R functions (bookmark this section)

These are functions you will use **throughout the semester**.

---

### Basic R functions

| Purpose | Function |
|------|---------|
| Create vector | `c()` |
| Mean | `mean()` |
| Median | `median()` |
| Standard deviation | `sd()` |
| Sample size | `length()` |
| Summary | `summary()` |
| View data | `head()` |

---

### Data wrangling (tidyverse)

| Task | Function |
|---|---|
| Select columns | `select()` |
| Filter rows | `filter()` |
| Create variables | `mutate()` |
| Sort data | `arrange()` |
| Group data | `group_by()` |
| Summarize data | `summarise()` |

---

### Data reshaping

| Task | Function |
|---|---|
| Wide → long | `pivot_longer()` |
| Long → wide | `pivot_wider()` |

---

### Visualization (preview)

| Task | Function |
|---|---|
| Create plot | `ggplot()` |
| Scatterplot | `geom_point()` |
| Line plot | `geom_line()` |
| Bar chart | `geom_col()` |
| Labels | `labs()` |

---

### Modeling (coming soon)

| Task | Function |
|---|---|
| Linear model | `lm()` |
| Logistic model | `glm()` |
| Model summary | `summary()` |
| Confidence intervals | `confint()` |

---

## Reproducibility expectations (review)

In this course:

- analysis and writing live together in Quarto  
- numbers should be generated by R, not typed  
- scripts and documents should be organized  
- work should render without errors  

Reproducibility is a professional skill, not just a technical one.

---

## Self-check (recommended)

Before moving forward, ask yourself:

1. Do I know when to use a script vs a Quarto document?  
2. Can I explain what rendering does?  
3. Can I write a sentence using inline R code?  
4. Do I recognize the core functions listed above?  

If not, revisit Weeks 1 and 2 before continuing.

---

## Looking ahead

Next week, you will begin **data visualization with ggplot2**, where you will apply everything you learned about Quarto, reproducibility, and tidy workflows.

This review lesson is your foundation — keep it bookmarked.
