---
title: "Lesson 10: Confidence Intervals & Uncertainty"
subtitle: "Interval Logic, Interpretation, and Communication"
---

## Overview

Confidence intervals are one of the most important tools in statistical inference. They allow us to move beyond a single estimate and instead communicate a **range of plausible values** for a population parameter.

In this lesson, we focus less on calculation and more on **interpretation, logic, and communication**, which are essential skills in public health and data science.

------------------------------------------------------------------------

## Learning Objectives

By the end of this lesson, students will be able to:

-   Explain the logic behind confidence intervals
-   Interpret confidence intervals in plain language
-   Distinguish correct vs incorrect interpretations
-   Communicate uncertainty effectively in public health contexts
-   Connect confidence intervals to hypothesis testing concepts

------------------------------------------------------------------------

## Assigned Readings

-   *OpenIntro Biostatistics*, Chapter 4\
-   *Statistical Inference via Data Science*, Chapter 9: Hypothesis Testing

------------------------------------------------------------------------

## What Is a Confidence Interval?

A confidence interval (CI) provides a **range of plausible values** for a population parameter.

Instead of reporting:

> "The mean BMI is 27.3"

we report:

> "The mean BMI is 27.3 (95% CI: 26.5 to 28.1)"

------------------------------------------------------------------------

## Interval Logic

Confidence intervals are built on three key components:

1.  **Estimate** (sample statistic)
2.  **Standard Error** (variability)
3.  **Critical Value** (confidence level)

Mathematically:

$$
\text{Estimate} \pm \text{Critical Value} \times SE
$$

------------------------------------------------------------------------

## Why Intervals Matter

A single number does not show uncertainty.

Confidence intervals:

-   show precision of estimates\
-   reflect sampling variability\
-   provide context for decision-making

------------------------------------------------------------------------

## Visualizing Uncertainty

```{r}
library(tidyverse)

set.seed(123)

# Simulate many confidence intervals
true_mean <- 50
population <- rnorm(10000, mean = true_mean, sd = 10)

ci_sim <- replicate(100, {
  samp <- sample(population, 50)
  m <- mean(samp)
  se <- sd(samp)/sqrt(50)
  lower <- m - 1.96 * se
  upper <- m + 1.96 * se
  c(lower, upper)
})

ci_df <- as.data.frame(t(ci_sim))
names(ci_df) <- c("lower", "upper")
ci_df$id <- 1:nrow(ci_df)
ci_df$contains_true <- ci_df$lower <= true_mean & ci_df$upper >= true_mean

head(ci_df)
```

------------------------------------------------------------------------

## Plotting Confidence Intervals

```{r}
ggplot(ci_df, aes(x = id, ymin = lower, ymax = upper, color = contains_true)) +
  geom_linerange() +
  geom_hline(yintercept = true_mean, linetype = "dashed") +
  labs(
    title = "Simulated 95% Confidence Intervals",
    x = "Sample",
    y = "Confidence Interval"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## Key Insight

-   Most intervals contain the true mean\
-   Some do not (about 5% for a 95% CI)

This is the meaning of "95% confidence"

------------------------------------------------------------------------

## Correct Interpretation

A correct interpretation is:

> We are 95% confident that the true population parameter lies within this interval.

------------------------------------------------------------------------

## What "95% Confidence" Really Means

If we repeated the sampling process many times:

-   95% of constructed intervals would contain the true parameter\
-   5% would miss it

------------------------------------------------------------------------

## Common Incorrect Interpretations

-   ❌ "There is a 95% probability the true value is in this interval"
-   ❌ "95% of individuals fall within this range"
-   ❌ "The parameter changes from sample to sample"

------------------------------------------------------------------------

## Interpreting Instead of Calculating

In practice, you will often be given confidence intervals rather than calculating them.

Example:

> Odds Ratio = 1.5 (95% CI: 1.2 to 1.9)

Interpretation:

-   The exposure is associated with higher odds of the outcome\
-   The interval does not include 1 → statistically significant\
-   The plausible effect size ranges from 1.2 to 1.9

------------------------------------------------------------------------

## Width of Confidence Intervals

The width of a confidence interval depends on:

-   Sample size (larger → narrower)\
-   Variability (higher → wider)\
-   Confidence level (higher → wider)

------------------------------------------------------------------------

## Demonstration: Sample Size Effect

```{r}
set.seed(456)

get_ci_width <- function(n) {
  samp <- sample(population, n)
  se <- sd(samp)/sqrt(n)
  width <- (1.96 * se) * 2
  width
}

tibble(
  sample_size = c(50, 200, 500),
  ci_width = c(get_ci_width(50), get_ci_width(200), get_ci_width(500))
)
```

------------------------------------------------------------------------

## Communicating Uncertainty

In public health, it is essential to communicate uncertainty clearly.

Instead of saying:

> "The intervention works"

Say:

> "The intervention is associated with improved outcomes (95% CI: X to Y), suggesting a likely positive effect, though uncertainty remains."

------------------------------------------------------------------------

## Public Health Example

Suppose:

-   Screening rate = 72%\
-   95% CI = 68% to 76%

Interpretation:

-   The true screening rate is likely between 68% and 76%\
-   The estimate is reasonably precise\
-   Policy decisions should consider this range

------------------------------------------------------------------------

## Confidence Intervals & Hypothesis Testing

Confidence intervals are closely related to hypothesis testing.

Key connection:

-   If the CI excludes the null value → statistically significant\
-   If the CI includes the null value → not statistically significant

Examples:

| Measure         | Null Value |
|-----------------|------------|
| Mean Difference | 0          |
| Risk Difference | 0          |
| Odds Ratio      | 1          |
| Risk Ratio      | 1          |

------------------------------------------------------------------------

## Example Interpretation

If:

-   Risk Ratio = 1.3 (95% CI: 0.9 to 1.8)

Then:

-   CI includes 1 → not statistically significant\
-   Evidence is inconclusive

------------------------------------------------------------------------

## Key Terms

| Term                     | Definition                       |
|--------------------------|----------------------------------|
| Confidence Interval      | Range of plausible values        |
| Margin of Error          | Distance from estimate to bounds |
| Precision                | Narrowness of interval           |
| Statistical Significance | CI excludes null value           |
| Null Value               | Value representing no effect     |

------------------------------------------------------------------------

## Practice Activity

1.  Given a confidence interval, interpret it in plain language\
2.  Determine whether it is statistically significant\
3.  Identify whether the interval is wide or narrow\
4.  Explain what that implies about precision

------------------------------------------------------------------------

## Reflection Questions

1.  Why are confidence intervals more informative than single estimates?\
2.  What does it mean when a confidence interval is wide?\
3.  How does sample size affect confidence intervals?\
4.  Why is correct interpretation important in public health?

------------------------------------------------------------------------

## Common Mistakes to Avoid

-   Treating CI as probability of parameter\
-   Ignoring width of interval\
-   Overstating certainty\
-   Misinterpreting statistical significance

------------------------------------------------------------------------

## Conclusion

Confidence intervals provide a powerful way to communicate uncertainty.

They allow us to:

-   quantify variability\
-   interpret results meaningfully\
-   connect estimation to decision-making

Understanding how to interpret and communicate confidence intervals is essential for evidence-based public health practice.

------------------------------------------------------------------------

## Looking Ahead

Next lesson:

-   Hypothesis testing in depth\
-   p-values\
-   Type I and Type II errors\
-   Decision-making frameworks

------------------------------------------------------------------------
