---
title: "Lesson 11: Hypothesis Testing"
subtitle: "Simulation-First Thinking, P-values, and Biomedical Interpretation"
---

## Overview

Hypothesis testing allows us to evaluate whether observed patterns in data are likely due to chance or reflect a real underlying relationship.

In this lesson, we take a **simulation-first approach** to hypothesis testing, emphasizing intuition, interpretation, and real-world biomedical framing rather than memorizing formulas.

------------------------------------------------------------------------

## Learning Objectives

By the end of this lesson, students will be able to:

-   Explain the logic of hypothesis testing using simulation
-   Define null and alternative hypotheses in public health contexts
-   Interpret p-values as measures of evidence strength
-   Avoid common misinterpretations of p-values
-   Apply hypothesis testing to real-world biomedical questions
-   Connect hypothesis testing to regression inference

------------------------------------------------------------------------

## Assigned Readings

-   *OpenIntro Biostatistics*, Chapter 5\
-   *Statistical Inference via Data Science*, Chapter 10: Inference for Regression

------------------------------------------------------------------------

## What Is Hypothesis Testing?

Hypothesis testing is a framework used to evaluate whether observed data are consistent with a specific claim about a population.

We compare:

-   what we **observed in our data**
-   what we would **expect under a null hypothesis**

------------------------------------------------------------------------

## Null and Alternative Hypotheses

The **null hypothesis (H₀)** represents no effect or no association.

The **alternative hypothesis (H₁ or Hₐ)** represents a meaningful effect or association.

------------------------------------------------------------------------

### Example (Public Health)

-   H₀: There is no difference in screening rates between groups\
-   Hₐ: There is a difference in screening rates between groups

------------------------------------------------------------------------

## Simulation-First Thinking

Instead of jumping straight to formulas, we ask:

> What would our data look like if the null hypothesis were true?

We then simulate many datasets under the null and compare our observed statistic to that distribution.

------------------------------------------------------------------------

## Example: Simulating a Null Distribution

```{r}
library(tidyverse)

set.seed(123)

# Create two groups with no real difference
group <- rep(c("A", "B"), each = 50)
outcome <- rnorm(100, mean = 50, sd = 10)

df <- tibble(group, outcome)

# Observed difference in means
obs_diff <- df |>
  group_by(group) |>
  summarise(mean_val = mean(outcome)) |>
  summarise(diff = diff(mean_val)) |>
  pull(diff)

obs_diff
```

------------------------------------------------------------------------

## Permutation Test (Simulation)

```{r}
set.seed(123)

null_dist <- replicate(1000, {
  shuffled <- df |>
    mutate(group = sample(group))

  shuffled |>
    group_by(group) |>
    summarise(mean_val = mean(outcome)) |>
    summarise(diff = diff(mean_val)) |>
    pull(diff)
})

null_df <- tibble(diff = null_dist)
```

------------------------------------------------------------------------

## Visualizing the Null Distribution

```{r}
ggplot(null_df, aes(x = diff)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = obs_diff, linetype = "dashed") +
  labs(
    title = "Null Distribution of Difference in Means",
    x = "Difference in Means",
    y = "Frequency"
  )
```

------------------------------------------------------------------------

## Interpreting the Simulation

-   The histogram shows what differences we would expect **if there were no real effect**
-   The dashed line shows our observed result
-   We assess how extreme our observed value is relative to the null distribution

------------------------------------------------------------------------

## What Is a P-value?

The **p-value** is the probability of observing a result as extreme (or more extreme) than what we observed, **assuming the null hypothesis is true**.

------------------------------------------------------------------------

## Calculating the P-value

```{r}
p_value <- mean(abs(null_df$diff) >= abs(obs_diff))
p_value
```

------------------------------------------------------------------------

## Interpreting the P-value

-   Small p-value → strong evidence against the null\
-   Large p-value → weak evidence against the null

------------------------------------------------------------------------

## P-values as Evidence Strength

| P-value   | Interpretation       |
|-----------|----------------------|
| \< 0.01   | Very strong evidence |
| 0.01–0.05 | Strong evidence      |
| 0.05–0.10 | Moderate evidence    |
| \> 0.10   | Weak evidence        |

------------------------------------------------------------------------

## What P-values Do NOT Mean

-   ❌ The probability that the null hypothesis is true\
-   ❌ The probability that the results are due to chance alone\
-   ❌ The size or importance of an effect

------------------------------------------------------------------------

## Decision-Making Framework

We often compare the p-value to a significance level (α), commonly 0.05.

-   If p-value \< 0.05 → reject H₀\
-   If p-value ≥ 0.05 → fail to reject H₀

------------------------------------------------------------------------

## Biomedical Framing

In public health and medicine, hypothesis testing answers questions like:

-   Does a treatment improve outcomes?
-   Is an exposure associated with disease?
-   Do disparities persist between populations?

------------------------------------------------------------------------

### Example Interpretation

> The intervention group had lower blood pressure than the control group (p = 0.03).

Interpretation:

-   There is evidence of a difference\
-   The observed result is unlikely under the null\
-   The effect may be meaningful, but magnitude should also be considered

------------------------------------------------------------------------

## Hypothesis Testing in Regression

Hypothesis testing is also used in regression models.

Example:

-   H₀: β₁ = 0 (no association between predictor and outcome)\
-   Hₐ: β₁ ≠ 0

------------------------------------------------------------------------

### Example in R

```{r}
model <- lm(outcome ~ group, data = df)
summary(model)
```

------------------------------------------------------------------------

## Interpreting Regression Output

-   The coefficient represents the estimated effect\
-   The p-value tests whether that effect is different from zero\
-   Confidence intervals provide a range of plausible values

------------------------------------------------------------------------

## Connecting to Confidence Intervals

-   If CI excludes 0 (or 1 for ratios) → statistically significant\
-   If CI includes 0 → not statistically significant

------------------------------------------------------------------------

## Key Terms

| Term                   | Definition                             |
|------------------------|----------------------------------------|
| Null Hypothesis        | No effect or association               |
| Alternative Hypothesis | Presence of effect                     |
| Test Statistic         | Measure of observed effect             |
| Null Distribution      | Distribution under H₀                  |
| P-value                | Probability of extreme result under H₀ |
| Significance Level     | Threshold for decision-making          |

------------------------------------------------------------------------

## Practice Activity

1.  Simulate a null distribution\
2.  Compute an observed statistic\
3.  Calculate a p-value\
4.  Interpret the result in plain language\
5.  Decide whether to reject the null

------------------------------------------------------------------------

## Reflection Questions

1.  Why do we simulate under the null hypothesis?\
2.  What does a small p-value tell us?\
3.  Why is statistical significance not the same as practical importance?\
4.  How does this apply to real public health decisions?

------------------------------------------------------------------------

## Common Mistakes to Avoid

-   Treating p-value as probability the null is true\
-   Ignoring effect size\
-   Over-relying on 0.05 cutoff\
-   Confusing statistical and clinical significance

------------------------------------------------------------------------

## Conclusion

Hypothesis testing provides a structured way to evaluate evidence.

A simulation-first approach helps build intuition by showing what results look like under the null hypothesis.

P-values quantify evidence strength, but must be interpreted carefully and alongside effect sizes and confidence intervals.

------------------------------------------------------------------------

## Looking Ahead

Next lesson:

-   Multiple testing\
-   Type I and Type II errors\
-   Power and sample size

------------------------------------------------------------------------
