---
title: "Lesson 11a: In-Class Activity: Hypothesis Testing Simulation"
---

## Activity Overview

In this activity, you will use simulation to understand hypothesis testing and p-values.

**Scenario:**\
A researcher wants to know whether a new intervention changes an outcome compared to a control group.

------------------------------------------------------------------------

## Step 1: Set Up the Data

```{r}
library(tidyverse)

set.seed(123)

group <- rep(c("Control", "Intervention"), each = 50)
outcome <- rnorm(100, mean = 50, sd = 10)

df <- tibble(group, outcome)

df
```

------------------------------------------------------------------------

## Step 2: Calculate Observed Difference

```{r}
obs_diff <- df |>
  group_by(group) |>
  summarise(mean_val = mean(outcome)) |>
  summarise(diff = diff(mean_val)) |>
  pull(diff)

obs_diff
```

------------------------------------------------------------------------

## Step 3: Simulate Under the Null

```{r}
set.seed(123)

null_dist <- replicate(1000, {
  shuffled <- df |>
    mutate(group = sample(group))

  shuffled |>
    group_by(group) |>
    summarise(mean_val = mean(outcome)) |>
    summarise(diff = diff(mean_val)) |>
    pull(diff)
})

null_df <- tibble(diff = null_dist)
```

------------------------------------------------------------------------

## Step 4: Visualize

```{r}
ggplot(null_df, aes(x = diff)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = obs_diff, linetype = "dashed") +
  labs(
    title = "Null Distribution",
    x = "Difference in Means",
    y = "Frequency"
  )
```

------------------------------------------------------------------------

## Step 5: Compute p-value

```{r}
p_value <- mean(abs(null_df$diff) >= abs(obs_diff))
p_value
```

------------------------------------------------------------------------

## Step 6: Interpretation Questions

-   Is the observed result extreme compared to the null?
-   Is the p-value small or large?
-   What does this say about evidence against the null?
-   Would you reject the null hypothesis?

------------------------------------------------------------------------

## Key Takeaway

A p-value tells us how surprising our data are under the null hypothesis.

Small p-value → strong evidence against the null\
Large p-value → weak evidence against the null
