---
title: "Project 1 Lesson: Exploratory Data Story with Gapminder"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose of Project 1

Project 1 is designed to help you practice the **core workflow** of this course before working with complex health datasets.

This project focuses on:

- working comfortably in R and Quarto  
- exploring a dataset using descriptive statistics  
- creating clear data visualizations  
- writing interpretations in plain public health language  
- using inline R code to report values reproducibly  

This is **not** a modeling project. The goal is exploration, clarity, and workflow.

---

## Why we use Gapminder for Project 1

The Gapminder dataset is:

- clean and well-documented  
- structured in tidy format  
- widely used in teaching data science  
- rich enough to tell meaningful public health stories  

Using Gapminder allows you to focus on **skills**, not data complexity. These same skills will transfer directly to NHANES later in the course.

---

## What you are expected to demonstrate

By the end of Project 1, you should be able to:

- work entirely in a Quarto document  
- load and explore a dataset  
- compute descriptive summaries  
- create at least two well-labeled figures  
- interpret patterns in writing  
- use inline code for reported values  

Effort, clarity, and reproducibility matter more than sophistication.

---

## Project setup

Create a new Quarto document named:

```
project1-lastname.qmd
```

At the top of your document, include a clear title and your name.

---

## Loading the data

Install packages if needed (only once):

```{r}
# install.packages("tidyverse")
# install.packages("gapminder")
```

Load packages each session:

```{r}
library(tidyverse)
library(gapminder)
```

Load the dataset:

```{r}
gapminder
```

---

## Choosing a focus

You will be assigned a country and you will be comparing all results to the **United States**.

Once selected, filter the dataset.

```{r}
country_df <- gapminder %>%
  filter(country == "United States")

country_df
```

Your entire project should focus on **one country and the clearly defined comparison**.

---

## Section 1: Introduction

In your Introduction section, briefly describe:

- the country you are examining  
- why it is of interest from a public health perspective  
- what outcomes you plan to explore  

This section should be written in complete sentences and should not include code.

---

## Section 1: Introduction

In your Introduction section, briefly describe:

- the country you are examining  
- why it is of interest from a public health perspective  
- what outcomes you plan to explore  

This section should be written in complete sentences and should not include code.

---

## Section 2: Descriptive exploration

Start by computing basic descriptive statistics.

```{r}
mean_lifeExp <- mean(country_df$lifeExp)
sd_lifeExp <- sd(country_df$lifeExp)
n_years <- length(unique(country_df$year))
```

Report values using **inline code**, not by typing numbers manually.

Example:

Across `r n_years` years of available data, the mean life expectancy was `r round(mean_lifeExp, 1)` years (SD = `r round(sd_lifeExp, 1)`).

---

## Section 3: Visualizations

You must include **at least two figures**.

### Required figure: trend over time

```{r}
ggplot(country_df, aes(x = year, y = lifeExp)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Life Expectancy Over Time",
    x = "Year",
    y = "Life Expectancy (years)"
  )
```

### Optional second figure (examples)

You may choose one of the following:

- GDP per capita vs life expectancy  
- population trends over time  
- comparison with another country  

Example:

```{r}
ggplot(country_df, 
       aes(
         x = gdpPercap, 
         y = lifeExp)
       ) +
  geom_point() +
  scale_x_log10() +
  labs(
    title = "Life Expectancy vs GDP per Capita",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)"
  )
```

---

## Writing interpretation paragraphs

Every figure must be followed by a short interpretation paragraph.

A strong interpretation includes:

- what the plot shows  
- notable trends or changes  
- a public health interpretation  
- a brief limitation  

Example structure:

- What do you observe?  
- When do changes occur?  
- Why might this matter for population health?  

Do not describe code or aesthetics. Focus on meaning.

---

## Section 4: Data quality and limitations

Include a short section discussing:

- missingness (if any)  
- time coverage of the data  
- what the data cannot tell us  

This helps build good scientific habits early.

---

## Section 5: Reflection

Conclude with a brief reflection addressing:

- what you learned from exploring the data  
- what surprised you  
- what questions you would explore next  

This section is about **thinking**, not technical skill.

---

## Formatting and reproducibility requirements

Your project must:

- be written entirely in Quarto  
- render without errors  
- use inline code for reported values  
- include labeled figures  
- avoid screenshots of plots  
- be clearly organized with headings
- have peer-reviewed sources (at least 8)

:::{.callout-warning}
Projects that do not render cleanly cannot be graded.
:::
---

## Submission instructions

Submit:

- the rendered document (word)  
- the `.qmd` source file  

Both are required.

---

## How this project prepares you for the rest of the course

Project 1 builds the foundation for:

- Project 2 (cleaning NHANES data)  
- Project 3 (modeling and inference)  
- Project 4 (final manuscript)  

The workflow you practice here will be reused all semester.

---

## Key takeaways

Project 1 is about learning how to explore data, write clearly, and work reproducibly in Quarto. Focus on effort, clarity, and honest interpretation. You are not expected to be an expert — you are expected to engage.

---

## Looking ahead

After Project 1, we will transition to **real U.S. health data** and begin preparing datasets for modeling and inference.
