---
title: "Supplemental Lesson: Creating Table 1, Table 2, and Code Chunks"
subtitle: "Alternative Method for `gtsummary()`"
---

## Overview

This lesson shows you how to: 

- Create **Table 1 (descriptive statistics)** 
- Create **Table 2 (regression results)** 
- Use **Quarto chunk options** to control code and output 
- Write cleaner, more impactful reports

------------------------------------------------------------------------

## Setup

```{r}
# install.packages(c("dplyr", "knitr", "broom"))
library(dplyr)
library(knitr)
library(broom)
```

------------------------------------------------------------------------

## Example Dataset

```{r}
data(starwars)

df <- starwars %>%
  select(sex, height, mass) %>%
  filter(!is.na(sex))
```

------------------------------------------------------------------------

## Table 1: Descriptive Statistics

### Simple Summary Table

```{r}
table1 <- df %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    mean_height = mean(height, na.rm = TRUE),
    mean_mass = mean(mass, na.rm = TRUE)
  )

kable(table1, caption = "Table 1. Descriptive Statistics by Sex")
```

------------------------------------------------------------------------

## Table 2: Regression Results

```{r}
model <- lm(mass ~ height, data = df)

table2 <- tidy(model)

kable(table2, digits = 3, caption = "Table 2. Linear Regression Results")
```

------------------------------------------------------------------------

## Quarto Tips: Controlling Code & Output

**REMEMBER** - The code goes INSIDE of the `{r}`. Make sure to use a comma after each new command.

### 1. Hide Code, Show Output

```{r, echo=FALSE}
summary(df$height)
```

The code: `echo=FALSE`

------------------------------------------------------------------------

### 2. Show Code, Hide Output

```{r, results='hide'}
summary(df$height)
```

The code: `results='hide'`

------------------------------------------------------------------------

### 3. Hide Both Code and Output

```{r, include=FALSE}
summary(df$height)
```

The code: `include=FALSE`

------------------------------------------------------------------------

### 4. Clean Tables (No Messages or Warnings)

```{r, message=FALSE, warning=FALSE}
kable(head(df), caption = "Clean Table Example")
```

The code: `message=FALSE, warning=FALSE`

------------------------------------------------------------------------

## Writing Tips for Quarto Reports

-   Keep text **short and clear**
-   Always **interpret your results**
-   Use **table captions** to explain what the reader is seeing
-   Avoid dumping raw output — format it into tables
-   Use headings to organize your report

### Example Interpretation

The results suggest that height is positively associated with mass, indicating that taller individuals tend to have higher body mass on average.

------------------------------------------------------------------------

## Key Takeaways

-   Use `kable()` for clean, simple tables
-   Always structure Table 1 (descriptive) and Table 2 (model results)
-   Control what your audience sees using chunk options
-   Focus on interpretation, not just output
