---
title: "Supplemental Lesson: Creating Table 1, Table 2, and References"
subtitle: "Applied Epidemiologic Reporting in Quarto"
---

## Overview

In this lesson, you will learn how to create **Table 1 (Descriptive Statistics)**, **Table 2 (Analytic Results)**, and properly format **references** in a reproducible Quarto workflow.

These are essential components of epidemiologic manuscripts and will be required for your final project.

------------------------------------------------------------------------

## Learning Objectives

By the end of this lesson, students will be able to:

-   Create a publication-style Table 1
-   Create a regression-based Table 2
-   Interpret tables in plain language
-   Export clean tables from R
-   Add and format references in Quarto

------------------------------------------------------------------------

## Required Packages

```{r}
library(tidyverse)
library(gtsummary)
library(gt)
library(broom)
```

------------------------------------------------------------------------

## Example Dataset

```{r}
library(NHANES)

df <- NHANES %>%
  select(Age, Gender, BMI, BPSysAve, Diabetes) %>%
  drop_na()
```

------------------------------------------------------------------------

## Table 1: Descriptive Statistics

Table 1 summarizes the characteristics of your study population.

### Step 1: Create Table 1

```{r}
table1 <- df %>%
  tbl_summary(
    by = Diabetes,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) %>%
  add_p()

table1
```

------------------------------------------------------------------------

## Interpretation of Table 1

-   Continuous variables: mean (SD)
-   Categorical variables: n (%)
-   p-values: compare groups

------------------------------------------------------------------------

## Export Table 1

```{r}
table1 %>%
  as_gt() %>%
  gt::gtsave("table1.html")
```

------------------------------------------------------------------------

## Table 2: Regression Results

Table 2 presents results from regression models.

### Step 1: Fit Model

```{r}
model <- lm(BMI ~ Age + Gender + Diabetes, data = df)
summary(model)
```

------------------------------------------------------------------------

### Step 2: Create Table 2

```{r}
table2 <- tbl_regression(model) %>%
  bold_labels()

table2
```

------------------------------------------------------------------------

## Interpretation of Table 2

-   Coefficients show direction and magnitude
-   Confidence intervals show uncertainty
-   p-values show statistical evidence

------------------------------------------------------------------------

## Optional: Logistic Regression

```{r}
log_model <- glm(Diabetes ~ Age + BMI + Gender,
                 data = df,
                 family = binomial)

tbl_regression(log_model, exponentiate = TRUE)
```

------------------------------------------------------------------------

## Export Table 2

```{r}
table2 %>%
  as_gt() %>%
  gt::gtsave("table2.html")
```

------------------------------------------------------------------------

## Writing About Tables

Example:

"The mean BMI was higher among individuals with diabetes compared to those without (p \< 0.05). In adjusted models, age and BMI were significantly associated with diabetes status."

------------------------------------------------------------------------

## Adding References in Quarto

### Step 1: Create a `.bib` File

Example `references.bib`:

```         
@article{nhanes,
  title={National Health and Nutrition Examination Survey},
  author={CDC},
  year={2023}
}
```

------------------------------------------------------------------------

### Step 2: Add to YAML

```         
bibliography: references.bib
csl: apa.csl
```

------------------------------------------------------------------------

### Step 3: Cite in Text

Example:

This dataset comes from NHANES [@nhanes].

------------------------------------------------------------------------

### Step 4: References Section

Add this at the end:

```         
## References
```

Quarto will automatically generate the reference list.

------------------------------------------------------------------------

## Key Takeaways

-   Table 1 describes your sample
-   Table 2 presents analytic results
-   Use `gtsummary` for clean outputs
-   Always interpret results in plain language
-   Use Quarto citation tools for references

------------------------------------------------------------------------

## Practice Activity

1.  Create Table 1 using your Project dataset\
2.  Create Table 2 using a regression model\
3.  Export both tables\
4.  Write 2–3 sentences interpreting each table\
5.  Add at least one reference using `.bib`

------------------------------------------------------------------------

## Conclusion

Tables and references are essential for scientific communication. Mastering these skills will help you produce clear, professional, and reproducible public health reports.

------------------------------------------------------------------------
